---
title: "Computing in 141Z"
subtitle: "R arithmetic syntax"
output: learnr::tutorial
tutorial:
  id: "Syntax"
  version: 0.1
runtime: shiny_prerendered
date: ""
tags: [arithmetic, second, third]
id: R-syntax
---

```{r include = FALSE}
# Note the context="server". That's so that the question_ui_initialize methods from etude will be registered.
library(mosaic)
library(mosaicCalc)
library(ggformula)
library(math141Z)
library(etude)
library(learnr)
library(gradethis)
library(submitr)
library(basket)
library(USAFAexercises)


learnr::tutorial_options(exercise.timelimit = 60,
                 exercise.checker = gradethis::grade_learnr)
```

```{r echo=FALSE}
submitr::login_controls()
```

```{r context="server", echo = FALSE}
options(tutorial.storage = "none")
vfun <- basket::check_valid
  #submitr::make_basic_validator(NULL, "hello")
storage_actions <- submitr::record_gs4("1w3fEld2rZlR_6FuzvkA-viOLBA5JdqD3xHl-LuLX3-Y", "statprep.annie@gmail.com", vfun)
  # submitr::record_local("./minimal_submissions.csv")
submitr::shiny_logic(input, output, session, vfun,
                     storage_actions)
```

## Syntax?

The word "syntax" means "the arrangement of words and phrases to create well-formed sentences in a language."  The language you will be using is R. By learning the syntax of R you will be able to make sense of new R commands without memorizing lots of details. 

The aspects of R syntax that you will learn in the next few tutorials are:

- basic arithmetic
- application of **functions** to **arguments**
- using **names**
- giving names to the objects you create,  also called "assignment."
- creating and using **tilde-expressions**, also called "formulas."
- the helpful use of blank space

This tutorial is about basic arithmetic notation in R.

##  Basic arithmetic

Almost all computer languages provide a way for writing numerals and doing arithmetic that is a close match to the tradition notation you learned in mathematics classes. So many of the following examples will be obvious to you. Still, there are little differences from time to time that are important to get right.

Traditional          | R                   | comments
--------------------:|:--------------------|:---------
$160$                |  `160`              | 
$1.6 \times 10^2$    | `1.6e2`             | The `e2` means $\times 10^2$. Note that there are no spaces in the R expression.
$1.6 \times 10^{-3}$ | `1.6e-3`            | The `e-3` means $\times 10^2$. Again, no spaces.
$15 + 3$             | `15 + 3`            |
$\frac{15}{3}$       | `15 / 3`            |
$15 \times 3$        | `15 * 3`            | You must always use the `*` for multiplication.
$15 \cdot 3$         | `15 * 3`            |
$15 - 3$             | `15 - 3`            |
$15^3$               | `15  ^  3`          | They `^` symbol is called a "caret." There are no superscripts in R. You use `^` instead.
$15^{3 + 2}$         | `15 ^ (3 + 2)`      | You need to use parenthesis to enforce the grouping of `3 + 2`.
$\frac{4 + 2}{1 + 5}$| `(4 + 2) / (1 + 5)` | Use parentheses to enforce the grouping.
$|3 - 5|$            | `abs(3 - 5)`        | Use the `abs()` function.
$\sqrt{15}$          | `sqrt(15)`          | Use the `sqrt()` function.  Note the parentheses.
$\sqrt[3]{15}$       | `15 ^ (1/3)`        | Recall that even in traditional notation, $\sqrt[3]{15}$  is  the same as $15^{\frac{1}{3}}$
$\sin \pi$           | `sin(pi)`           | Note that the number $\pi$ is given the *name* `pi` in R. Also note the parentheses. 
$\ln 20$             | `log(20)`           | The so-called "natural" logarithm.
$\log_{10}20$        | `log(20, base = 10)`
$e^{3}$              | `exp(3)`            | Don't confuse the mathematical number $e$ (which is called "Euler's number") with the use of `e` in scientific notation in R, for instance `1.6e-3`.

## Practice 

These are drawn from Question 7 on p. 13 of MMAC textbook.

`r etude::include_etude("Exercises/Computing/bee-write-closet.Rmd", "**MMAC 1.1.7a**: ", package = "USAFAexercises")`

`r etude::include_etude("Exercises/Computing/bee-write-closet2.Rmd", "**MMAC 1.1.7b**: ", package = "USAFAexercises")`

`r etude::include_etude("Exercises/Computing/bee-write-closet3.Rmd", "**MMAC 1.1.7c**: ", package = "USAFAexercises")`

`r etude::include_etude("Exercises/Computing/bee-write-closet4.Rmd", "**MMAC 1.1.7d**: ", package = "USAFAexercises")`

`r etude::include_etude("Exercises/Computing/bee-write-closet5.Rmd", "**MMAC 1.1.7e**: ", package = "USAFAexercises")`

`r etude::include_etude("Exercises/Computing/bee-write-closet6.Rmd", "**MMAC 1.1.7f**: ", package = "USAFAexercises")`

-----

`r etude::include_etude("Exercises/Computing/panda-dig-book.Rmd", "**Bonus**: ", package = "USAFAexercises")`




