---
title: "Computing in 141Z"
subtitle: "MMAC Chapter 1"
output: learnr::tutorial
tutorial:
  id: "MMAC-Ch-1"
  version: 0.1
runtime: shiny_prerendered
date: ""
tags: [MMAC, second, third]
id: R-syntax
---

```{r include= FALSE}
# In the document context to handle the pre-rendered chunks
library(mosaic)
library(mosaicCalc)
```

```{r include = FALSE, context="server"}
# Note the context="server". That's so that the question_ui_initialize methods from etude will be registered.
library(mosaic)
library(mosaicCalc)
library(ggformula)
library(math141Z)
library(etude)
library(learnr)
library(gradethis)
library(submitr)
library(basket)
library(USAFAexercises)
etude::show_answers(FALSE)


learnr::tutorial_options(exercise.timelimit = 60,
                 exercise.checker = gradethis::grade_learnr)
```

```{r echo=FALSE}
submitr::login_controls()
```

```{r context="server", echo = FALSE}
options(tutorial.storage = "none")
vfun <- basket::check_valid
  #submitr::make_basic_validator(NULL, "hello")
storage_actions <- submitr::record_gs4("1w3fEld2rZlR_6FuzvkA-viOLBA5JdqD3xHl-LuLX3-Y", "statprep.annie@gmail.com", vfun)
  # submitr::record_local("./minimal_submissions.csv")
submitr::shiny_logic(input, output, session, vfun,
                     storage_actions)
```


##  MMAC Chapter 1

This tutorial covers the R commands used in  Chapter 1 of Kilty and McAllister's *Mathematical Modeling and Applied Calculus*.  You can ignore the "Working in RStudio" sections in the book itself. 

In this tutorial, you  will learn

* How to implement a mathematical formula as an R function.
* How to draw graphs of functions with one input.
* How to draw graphs of functions with two inputs.
* How to solve simultaneous equations graphically.
* How to direct the computer to solve simultaneous equations numerically.
* How to show inverse functions graphically. (flip_coord)
* The R functions implementing exponential, logarithmic, sine and cosine relationships.
* How to make graphics with semi-log and log-log axes.

The new R functions you will see  are:

- `makeFun()` -- turns a tilde-expression into a function
- `slice_plot()` -- graphs a function with one input
- `domain()` -- sets the extent of the $x$ axis (the "domain")
- `c()` -- collects its argument together into a <span title="A vector is a collection of two or more numbers." style="text-decoration: underline">vector</span>.
- `contour_plot()` -- graphs a function with two inputs
- `findZeros()` -- finds solutions numerically
- `gf_refine()` -- adds selected refinements to plots
- `coord_flip()` -- flips the $x$ and $y$ axes in a plot.  
- `exp()`, `log()`,  `sin()`, `cos()` -- the R implementation of the <span title="eponymous means 'the thing that gives its name to something'" style="text-decoration: underline">eponymous</span> mathematical functions.

Make sure to review the **R Syntax** tutorial [IN DRAFT: GIVE LINK] before starting this tutorial. 

## Formulas to functions

A formula is a mathematical expression that describes an arithmetic calculation. For example, $m c^2$ is a formula introduced by Albert Einstein to describe the energy of a particle as a function of mass and the speed of light.

Math textbooks love to write functions using $x$ and $y$, for instance,

$$y =  5 + 2 x . $$

In this traditional notation, the formula is on the right-hand side of `=`. The $y$ on the left-hand side of `=` is more or less meaningless. There is nothing explicit in the notation to say what the input(s) are to the function.

In the MMAC textbook, they use this traditional notation often. In our own notes, we will use a better mathematical notation:

$$f(x) \equiv 5 + 2 x .$$
You may wonder what makes this notation better. Clearly, it is more verbose. But that provides some advantages:

- the function is given a name, in this case $f$.
- the argument(s) to the function are explicitly identified by the names included in the parentheses.
- using $\equiv$ rather than $=$ reminds us that this is a definition of $f$, rather than a statement that the left side happens to be equal to the right side.

Consider Einstein's energy formula. The function based on  this formula could be written:

$$E(m)  \equiv m c^2 .$$
Notice  that $E$ is simply a function of $m$. That's because the speed of light is *constant*, roughly 300,000km per second.

Let's implement $E(m)$ as an R function. We'll do this with a tilde expression that specifies not only the formula ($m c^2$) but which of the symbols in the formula are variables and which not. The tilde expression for Einstein's energy function is

    m * c^2 ~ m

To transform the tilde expression into an R function, use `makeFun()`, like this:

```{r}
E <- makeFun(m * c^2 ~ m)
```

You might wonder how the function `E()` knows the value of `c`. That's a topic for Chapter 2.

Finally, a small reminder. Traditional mathematical notation uses very extensively one-letter names, like $E$ or $x$ or $y$.  With computer notation, you can often be more effective by selecting names that remind you of what the thing you're creating is for. For instance, perhaps better to have created a function `mass_energy()` rather than `E()`, like this:

```{r}
mass_energy <- makeFun(mass * speed_of_light^2 ~ mass)
```

### Practice

`r etude::include_etude("Exercises/Computing/child-leave-coat.Rmd", "**Practice 1**: ", package = "USAFAexercises")`

`r etude::include_etude("Exercises/Computing/doe-stand-magnet.Rmd", "**MMAC 1.1.81**: ", package = "USAFAexercises")`

## Graphing functions of one variable

[This section replaces the "Working in RStudio" box on p. 14 of the MMAC book.]

The `slice_plot()` function makes graphs of mathematical functions with a single input. As you know, it's conventional to draw the graph with the function input on the horizontal axis and the function output on the vertical axis.

When making a graph, you need to choose what region of the number line the horizontal axis should cover. This region is called the *domain* of the graphic. Many mathematical functions have function domains that cover the whole number line (that is, $-\infty$ to $+\infty$) or the positive half of the number line (that is, 0 to $+\infty$). But even though you can legitimately evaluate the mathematical function at any point in its function domain, it's not possible to draw a graph that runs to $-\infty$ or $+\infty$. (Imagine how big the computer screen would have to be! Even a screen the size of the galaxy wouldn't do it.) So, you have to make choices of the part of the function domain that is to be included in the graphic domain. Typically this choice is based on context: the region of interest.

The `domain()` function is used within `slice_plot()` to set the graphic domain. Here's an example:

```{r echo=TRUE}
slice_plot(tanh(x) ~ x, domain(x = c(-2,4)))
```

Just FYI, the mathematical $\tanh()$ is called the "hyperbolic tangent" or, for short, pronounced "tanch" which rhymes with "branch." It is an example of a "sigmoidal function," so called because it has a double bend similar to the one in the letter "S."

Let's take apart the how `slice_plot()` is used.

- `slice_plot()` takes two arguments. The first is a tilde expression of exactly the same sort as used in `makeFun()`. The second is the domain, as produced by the `domain()` function.

- `domain()` itself takes arguments. Here, since there is only one input variable in the function, there is a single argument: a named argument with the same name as used on the right-hand side of the tilde expression. The *value* of that named argument is a pair of numbers: one specifying the left side of the graphic domain, the other specifying the right side. In our example, the argument is `x = c(-2, 4)`, setting the graphic domain to be $-2 \leq x \leq 4$. (Domains are often written in a different notation. In that notation, the domain would be written $x \in [-2, 4]$.)

- `c()` does something simple but essential. It collects the two numbers--here $-2$ and $4$--into a single package. That two-number package becomes the argument `x`, allowing you to specify both ends of the graphic domain at once. Some people prefer to use the `range()` function, which is another way of producing a package of two numbers. Either will work fine.

### Practice

`r etude::include_etude("Exercises/Computing/owl-choose-roof.Rmd", "**MMAC Question 1.1.8**: ", package = "USAFAexercises")`

## Functions with two or more inputs

You can use `makeFun()` to turn mathematical formulas with two or more inputs into functions. It's all a matter of the tilde expression. On the right side of the tilde, list the names of all the inputs, separating them by  `+`. Like this example, from Example 1.2.10 of the MMAC textbook for a function

$$f(x, y, z) \equiv x^2 + 7y - z .$$

```{r echo=TRUE}
f <- makeFun(x^2 + 7*y - z ~ x + y + z)
```

Remember that on the right side of the tilde, `x + y  + z` here, the `+` symbol means simply "and" or "as well as," not mathematical addition. Some people prefer to use `&` rather than `+` to avoid giving the wrong impression. That's fine.

On the left side of the tilde, `x^2 + 7*y - z` here, the symbols `+`, `^`, `*`, and `-` do have exactly their ordinary arithmetic meaning.

Plotting a function of two variables can be done in several ways. In these note (and in the MMAC book), we prefer to use a contour-plot format. Contour plots are produced by `contour_plot()` which takes two arguments:

1. A tilde expression with two variables on the right of the tilde, e.g. `x + y` or `capital + labor`.
2. A domain produced by `domain()` used with two arguments, one for each of the inputs listed in the tilde expression.

For instance, consider the function 

$$g(x, y) \equiv 5 - x - y/2$$
and suppose we want to plot it over the domain $-5 \leq x \leq 5$ and $-7 \leq y \leq 7$. You can accomplish this with

```{r echo=TRUE}
contour_plot(
  5 - x - y/2 ~ x + y,
  domain(x = c(-5,5), y = c(-7, 7))
  )
```

Since the tilde expression and the `domain()` specification are typographically long, for clarity I've put each of the two arguments to `contour_plot()` on it's own line. Notice that the two arguments are separated by a comma in the usual fashion. Notice as well that the closing function-application  parenthesis for `contour_plot()` is on its own line. Such formatting is helpful for the human reader and works well with R. 

Recall that in a contour plot, each individual contour shows the set of inputs at which the *output* of the function is a certain level that's labeled on the contour. The levels displayed in the plot are, by default, selected by `contour_plot()` to give a fair display of the range of function outputs over the graphic domain of the plot. To keep the plot from becoming overburdened with labels, `contour_plot()` will omit some of the labels. So in the example above, the contours are for function outputs of -2, 0, 2, 4, 6, 8, 10, and 12. But there are labels only for -2, 2, 6, and 10.

Sometimes you will want to choose for yourself the output levels at which a contour will be drawn and label every contour rather than skipping. You can accomplish this by giving additional named arguments to `contour_plot()`, for instance:

```{r echo=TRUE}
contour_plot(
  5 - x - y/2 ~ x + y,
  domain(x = c(-5,5), y = c(-7, 7)),
  contours_at = c(0, 3, 6, 9, 12),
  skip = 0
  )
```

By default, `contour_plot()` uses two modes to display the function: the contours as well as a color fill where the color indicates the output value at each point. This works pretty well on a computer display, but occasionally you may want to turn off the fill. To accomplish this, use the `filled = FALSE` named argument.

```{r echo=TRUE}
contour_plot(
  5 - x - y/2 ~ x + y,
  domain(x = c(-5,5), y = c(-7, 7)),
  contours_at = c(0, 3, 6, 9, 12),
  skip = 0,
  filled = FALSE
  )
```


### Practice

`r etude::include_etude("Exercises/Computing/giraffe-sing-magnet.Rmd", "**Practice 1**: ", package = "USAFAexercises")`

## Surface plots

Occasionally, it's helpful to display a function of two variables as if it were a three-dimensional object with the two inputs constituting two of the dimensions and the output represented by the third dimension. When the function is smooth, the output will a surface suspended above the input axes, analogous to an awning suspended over the two dimensions of a backyard patio.

Surface plots can be cool looking, but they can be hard to interpret quantitatively and it's practically impossible to compare two or more functions in surface plots. That's why we mostly use contour plots in displaying functions of two variables.

To make a surface plot, use the  `interactive_plot()` function, which takes a tilde expression and a graphic domain as arguments, just like `contour_plot()`.

```{r}
set.seed(102)
tmp <- mosaic::rfun( ~ latitude + longitude, n = 3)
terrain <- makeFun(tmp(latitude - 38, longitude + 104) ~ latitude + longitude)
```

```{r echo=TRUE}
interactive_plot(
  terrain(latitude, longitude) ~ latitude + longitude,
  domain(latitude = c(33, 43), longitude = c(-99, -106 ))
  )
```
Surface plots are much easier to interpret if you can rotate them. Use your mouse or trackpad to do this.

