---
title: "Computing in 141Z"
subtitle: "Syntax of functions application"
output: learnr::tutorial
tutorial:
  id: "Syntax"
  version: 0.1
runtime: shiny_prerendered
date: ""
tags: [arithmetic, second, third]
id: R-syntax
---

```{r include = FALSE}
# Note the context="server". That's so that the question_ui_initialize methods from etude will be registered.
library(mosaic)
library(mosaicCalc)
library(ggformula)
library(math141Z)
library(etude)
library(learnr)
library(gradethis)
library(submitr)
library(basket)
library(USAFAexercises)


learnr::tutorial_options(exercise.timelimit = 60,
                 exercise.checker = gradethis::grade_learnr)
```

```{r echo=FALSE}
submitr::login_controls()
```

```{r context="server", echo = FALSE}
options(tutorial.storage = "none")
vfun <- basket::check_valid
  #submitr::make_basic_validator(NULL, "hello")
storage_actions <- submitr::record_gs4("1w3fEld2rZlR_6FuzvkA-viOLBA5JdqD3xHl-LuLX3-Y", "statprep.annie@gmail.com", vfun)
  # submitr::record_local("./minimal_submissions.csv")
submitr::shiny_logic(input, output, session, vfun,
                     storage_actions)
```

## Function evaluation

In this tutorial, we going to look how to write an R command to *evaluate* a function. For example, you have already seen the `sqrt()` function. The command that follows evaluates the square-root function with the input 5.

```{r}
sqrt(5)
```



This is such an important concept in computing that there are many equivalent ways of saying it. Just for fun, here are a few:

- "Invoke a function."
- "Execute a function."
- Old timers (born pre-1970): 
    - "Call a function."
    - "Run a function."
- Grandparents (born pre-1960): "Call a subroutine."
- "Apply a function to arguments." -- The longest, but the most technically correct. The word "argument" is the proper way of referring to an input to a function.


## Spotting functions in commands

Consider this complicated-looking mathematical expression:

$$\sqrt{\sin^2 \frac{\pi}{5} + \cos^2 \frac{\pi}{5}} .$$

In R, this expression might be written like this:

```{r}
angle <- pi / 5
sqrt(sin(angle)^2 + cos(angle)^2)
```


- basic arithmetic
- application of **functions** to **arguments**
- using **names**
- giving names to the objects you create,  also called "assignment."
- creating and using **tilde-expressions**, also called "formulas."
- the helpful use of blank space




## Practice 

These are drawn from Question 7 on p. 13 of MMAC textbook.

`r etude::include_etude("Exercises/Computing/bee-write-closet.Rmd", "**MMAC 1.1.7a**: ", package = "USAFAexercises")`




