---
title: "Computing in 141Z"
subtitle: "Starting with R"
output: learnr::tutorial
tutorial:
  id: "Orientation"
  version: 0.1
runtime: shiny_prerendered
date: ""
tags: [first, second, third]
id: starting_with_R
---

```{r include = FALSE}
# Note the context="server". That's so that the question_ui_initialize methods from etude will be registered.
library(mosaic)
library(mosaicCalc)
library(ggformula)
library(math141Z)
library(etude)
library(learnr)
library(gradethis)
library(submitr)
library(basket)
library(USAFAexercises)


learnr::tutorial_options(exercise.timelimit = 60,
                 exercise.checker = gradethis::grade_learnr)
```

```{r echo=FALSE}
submitr::login_controls()
```

```{r context="server", echo = FALSE}
options(tutorial.storage = "none")
vfun <- basket::check_valid
  #submitr::make_basic_validator(NULL, "hello")
storage_actions <- submitr::record_gs4("1w3fEld2rZlR_6FuzvkA-viOLBA5JdqD3xHl-LuLX3-Y", "statprep.annie@gmail.com", vfun)
  # submitr::record_local("./minimal_submissions.csv")
submitr::shiny_logic(input, output, session, vfun,
                     storage_actions)
```

## Welcome

Welcome to the Math 141Z tutorials on using R for calculus.

These computing tutorials **replace** the "Working in RStudio" sections in the Kilty and McAllister *Mathematical Modeling and Applied Calculus* textbook. You can ignore those sections.

Each computing tutorial will have a layout similar to this document. For most browser setups, there are two columns in a tutorial. The left column gives the document title and a navigable table of contents. 

The right column shows one page of the tutorial. There are typically several pages in a computing tutorial.

Every page in the tutorial has a login section. You are to use the computing ID [IN DRAFT: We need to establish the system for this.] and password that have been provided to you. When you see the "Login SUCCESSFUL" message, you are ready to go. 

The login section appears at the top of every page so that you can confirm that you are correctly logged in or fix the situation if you are not. 

The content of the page is displayed after the login section. Some pages, like this one, will consist only of text narrative. Other pages will have a mix of material, as you will see.

The page ends with one or more buttons to move to the next or previous page. You could do the same thing using the table of contents. 

## Explanatory material

Many pages will contain explanatory material and notes. Not  surprisingly, some of this material will consist of illustrative  computer commands and the output generated by them. For the sake of acquainting you with the format of such illustrative commands, here is an example. For the moment, don't worry about understanding the content of the command.

```{r}
gf_point(kepler.radius ~ time, data = Kepler) %>%
  slice_plot(1.52 + 0.142 * sin(2 * pi  * time/687 - 300.3) ~ time)
```

The computer command appears in a grey box. The output from the command (if any) appears below. (The graph shows the distance of Mars from the Sun based on observations by Tycho Brahe (1546-1601) and analyzed by Johannes Kepler (1571-1630).)

## Code boxes

The tutorials include facilities for you to write your own computer commands and evaluate them. This is done in "code boxes" such as the one below.

```{r orientation-1, exercise=TRUE}
sqrt(47)
```

You press "Run code"  to evaluate the code in the box.

Often the code boxes will be pre-populated with text that you will edit in order to customize for some purpose. The following code box has a simple exercise:

There's a mistake in the code box below. Fix it in order to calculate $\sqrt{477}$.


```{r orientation-E1, exercise=TRUE, exercise.cap="Sandbox", exercise.setup="orientation-E1-setup", exercise.eval=FALSE, exercise.lines=2, exercise.completion=FALSE, exercise.startover=TRUE, exercise.diagnostics=TRUE}
sqt(47)
```

```{r orientation-E1-check}
gradethis::grade_result(
  pass_if( ~ .result == sqrt(477)),
  fail_if( ~ .result == sqrt(47), message = "We asked for the square root of 477. You need to do some more editing."),
  fail_if( ~ TRUE, message = "Keep trying!")
)
```

## Multiple choice questions

Tutorials may also include multiple-choice questions. Here's a silly example.  To see how things work, give an *incorrect* answer and then a correct one.

```{r document_25268ee-Qinline1, echo=FALSE}
etude::choose_one(
  prompt = "Which  is a vowel?",
  list("+A+" = "The other vowels are E, I, O, U,  and sometimes Y.", "B" = "B is a consonant, not a vowel.", "C" = "C is a consonant."),
  random_answer_order = FALSE
)
```

##  Text answers

```{r orientation-Essay1, echo=FALSE}
# indicate correct choices with +_+ in the name of the list item.
etude::essay_response(
  prompt = "Have you used computers before in a math class? If so, say what language was used.")


